Go Implementation of Up-To Techniques for Equivalence of Weighted Languages

### 2020-09-28 paper/paper.tex
intersection of linear subspaces
working on intersection and paper

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage{amsfonts}

@@

\DeclarePairedDelimiter\sqmap{\llbracket}{\rrbracket}

% matrix commands
\newcommand{\vvec}[2]{
	\begin{bmatrix}
		#1 \\ #2
	\end{bmatrix}
}
\newcommand{\vvvec}[3]{
	\begin{bmatrix}
		#1 \\ #2 \\ #3
	\end{bmatrix}
}

@@

\begin{abstract}
Weighted automata generalize non-deterministic automata by adding
a quantity expressing the weight (or probability) of the execution of each transition.
In this work we propose an implementation of two algorithms for computing language 
equivalence in finite state weighted automata (WAs). The first, a
linear partition refinement algorithm, calculates the largest linear weighted bisimulation
for any given LWA (Linear Weighted Automata) through an iterative method, 
the second algorithm checks the language equivalence 
of two vectors (states) for a given weighted automata by using an additional
data structure representing a congruence relationship between states.
We then compare the two algorithms results on randomly generated automata samples,
to verify their correctness and provide some runtime statistics.

@@

\begin{defn}
  A \textit{weighted automaton} over a field $\K$ and an alphabet $A$ is a triple 
  $(X,o,t)$ such that $X$ is a finite set of states, $t = (t_a : X \to K)_{a \in A}$
  is a set of transition functions indexed over the symbols of the alphabet and $o : X 
  \to \K$ is the output function. The transition functions $t_a$ are represented as
  $\K^{X \times X}$ matrices. $o \in \K^{1 \times X}$ is represented as a row vector.
  $t_a(v)$ denotes the vector obtained by multiplying the matrix $t_a$ by the column 
  vector $v  \in \K^{X \times 1}$. $o(v)$ denotes the scalar $s \in \K$ obtained by 
  multiplying the row vector $o$ by the column vector $v \in \K^{X \times 1}$.
  $A^*$ is the set of all words over $A$. $\epsilon$ is the empty word and $aw$ is the
  concatenation of a letter $a$ to the word $w \in A^*$.
  A weighted language is a function $\psi: A^* \to \K$.
  A function mapping each state vector into its 
  accepted language, $\sqmap{\cdot}: \K^X \to K^{A^*}$ is defined as follows for every weighted automaton:

@@

  Two vectors $v_1, v_2 \in \K^{X\times 1}$ are called language equivalent, denoted with $v_1 \sim v_2 \iff \sqmap
  {v_1} = \sqmap{v_2}$. One can extend the notion of language equivalence to states rather
  than for vectors by assigning to each state $x \in X$ the corresponding unit vector 
  $e_x \in \K^X$. When given an initial state $i$ for a weighted automaton, the language 
  of the automaton can be defined as $\sqmap{i}$.
\end{defn}

@@

@@

\subsection{Headings: second level}

@@

\subsection{Lists}
\begin{itemize}
\item Lorem ipsum dolor sit amet
\item consectetur adipiscing elit. 
\item Aliquam dignissim blandit est, in dictum tortor gravida eget. In ac rutrum magna.
\end{itemize}

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[table]{xcolor}
\usepackage{stmaryrd}
\usepackage{amsfonts}

% Code with syntax highlighting
\usepackage{fancyvrb}
\usepackage{listings}
\usepackage{textcomp}
\usepackage{placeins}

@@

\DeclarePairedDelimiter\sqmap{\llbracket}{\rrbracket}
\DeclarePairedDelimiter\spanset{\langle}{\rangle}

\newcommand{\diag}[1]{\text{diag}\left(#1\right)}
\newcommand{\mspan}[1]{\text{span}\left(#1\right)}
\newcommand{\dom}{\text{dom}}
\newcommand{\mker}[1]{\text{ker}\left(#1\right)}
\newcommand{\mrank}[1]{\text{rank}\left(#1\right)}

% matrix commands
\newcommand{\vvec}[2]{
	\begin{pmatrix}
		#1 \\ #2
	\end{pmatrix}
}
\newcommand{\vvvec}[3]{
	\begin{pmatrix}
		#1 \\ #2 \\ #3
	\end{pmatrix}
}
\newcommand{\vvvvec}[4]{
	\begin{pmatrix}
		#1 \\ #2 \\ #3 \\ #4
	\end{pmatrix}
}

% Code Listings
\definecolor{vgreen}{RGB}{104,180,104}
\definecolor{vblue}{RGB}{49,49,255}
\definecolor{vorange}{RGB}{255,143,102}
\lstset{
	basicstyle=\ttfamily,
	columns=fullflexible,
	keepspaces=true,
}

@@

\begin{abstract}
Weighted automata generalize non-deterministic automata by adding
a quantity expressing the weight (or probability) of the execution of each transition.
In this work we propose an implementation of two algorithms for computing language 
equivalence in finite state weighted automata (WAs). The first, a
linear partition refinement algorithm, calculates the largest linear weighted bisimulation
for any given LWA (Linear Weighted Automaton) through an iterative method, 
the second algorithm checks the language equivalence 
of two vectors (states) for a given weighted automata by using an additional
data structure representing a congruence relationship between states.
We then compare the two algorithms results to verify their correctness
on randomly generated automata samples, and provide some runtime statistics.

@@

\begin{defn}
  A \textit{weighted automaton} over a field $\K$ and an alphabet $A$ is a triple 
  $(X,o,t)$ such that $X$ is a finite set of states, $t = (t_a : X \to K)_{a \in A}$
  is a set of transition functions indexed over the symbols of the alphabet $A$ and $o : X 
  \to \K$ is the output function. The transition functions $t_a$ are represented as
  $\K^{X \times X}$ matrices. $o \in \K^{1 \times X}$ is represented as a row vector.
  $t_a(v)$ denotes the vector obtained by multiplying the matrix $t_a$ by the column 
  vector $v  \in \K^{X \times 1}$. $o(v)$ denotes the scalar $s \in \K$ obtained by 
  dot product of the row vector $o$ with $v \in \K^{X \times 1}$.
  $A^*$ is the set of all words over $A$. $\epsilon$ is the empty word and $aw$ is the
  concatenation of a symbol $a$ to the word $w \in A^*$.
  A weighted language is a function $\psi: A^* \to \K$.
  A function mapping each state vector into its 
  accepted language, $\sqmap{\cdot}: \K^X \to K^{A^*}$ is defined as follows for every weighted automaton:

@@

  Two vectors $v_1, v_2 \in \K^{X\times 1}$ are called language equivalent, denoted with 
  $v_1 \sim v_2 $ if and only if $ \sqmap
  {v_1} = \sqmap{v_2}$. One can extend the notion of language equivalence to states rather
  than for vectors by assigning to each state $x \in X$ the corresponding unit vector 
  $e_x \in \K^X$. When given an initial state $i$ for a weighted automaton, the language 
  of the automaton can be defined as $\sqmap{i}$.
\end{defn}

@@

\section{Algorithms}

\subsection{Forward Partition Refinement Algorithm for the Largest Weighted Bisimulation}
\label{sec:algo2}

The algorithm is defined by the iterative method:
\begin{eqnarray}
  R_0 = \mker{o}, & \quad & R_{i+1} = R_i \cap \bigcap_{a \in A} t_a(R_i)^{-1} \label{fwd} 
\end{eqnarray}
The algorithm stops when $R_{j+1} = R_j$. Proof is available in section 4.1 of \cite{BONCHI201277}

\subsection{Backwards Partition Refinement Algorithm for the Largest Weighted Bisimulation}
\label{sec:algo2}

The algorithm is defined by the iterative method:
\begin{eqnarray}
  R_0 = \mker{o}^0, & \quad & R_{i+1} = R_i + \sum_{a \in A} t_a(R_i)^t \label{back} 
\end{eqnarray}
The algorithm stops when $R_{j+1} = R_j$. Proof is available in section 4.2 of \cite{BONCHI201277}

@@

\subsection{Input file format}

\subsection{Calculating the intersection of linear subspaces}

To compute algorithm (\ref{fwd}), we need an efficient way to calculate a basis 
of the intersection of an arbitrary number of linear subspaces of $\R^n$, given their 
spanning sets.

\begin{defn}
  \textbf{Algorithm for a basis of the intersection of subspaces} \\
  Defined in file \texttt{lin/intersect.go}.
  Let's consider two linear subspaces of 
  $\R^n$ defined with their spanning sets of column vectors: 
  $U = \spanset{u_1, u_2, \hdots, u_p}$ 
  and $W = \spanset{w_1, w_2, \hdots, w_k}$.
  We create the block matrix $A \in \R^{n \times (p+k)}$ defined as follows:

  \begin{equation*}
    \begin{aligned}
      a_{ij} = \begin{cases}
        (u_j)_i & \text{for  } j = 1, \hdots, p \quad i = 1, \hdots, n \\
        (w_{j-p})_i & \text{for  } j = p+1, \hdots, p+k \quad i = 1, \hdots, n \\
      \end{cases}
    \end{aligned}
  \end{equation*}

  The matrix will have the form
  \begin{equation*}
    A = \begin{bmatrix}
      \vvvec{}{u_1}{\,}, \hdots \vvvec{}{u_p}{\,}, 
      \vvvec{}{w_1}{\,}, \hdots \vvvec{}{w_k}{\,} 
    \end{bmatrix}
  \end{equation*}

  It follows that $\mker{A}$ is a basis for $U \cap W$. This can be done with an 
  arbitrary finite number of subspaces of $\R^n$, as long as their spanning set is 
  \textit{finite}.
\end{defn}

\begin{defn}
  \textbf{Algorithm for computing the nullspace} \\
  The algorithm implementation can be found in file \texttt{lin/nullspace.go}; it is
  adapted from \cite{scipy/ranknullspace}. 
  Let's consider the singular value decomposition of a matrix $A \in \R^{m \times n}$:

  \begin{equation*}
    \begin{aligned}
      A = U \Sigma V^T & \quad & \Sigma = \diag{\sigma_1, \sigma_2, \hdots, \sigma_{\min(m, n)}  } 
       & \quad &  U \in \R^{m \times m} & \quad & V \in \R^{n \times n}
    \end{aligned}
  \end{equation*}

  Where $V$ and $U$ are orthogonal and the singular values are ordered: $\sigma_1 \geq \sigma_2 \geq \hdots \geq \sigma_{\min(m,n)} \geq 0$.
  It follows that $\mrank{A}$ is equal to the number of nonzero singular values and
  a basis of the (right) null space of $A$ is the spanning set of the columns of V
  corresponding to singular values equal to zero. 
\end{defn}

\begin{exmp}
  The actual Go implementation is quite verbose, we can show a shorter Python 
  implementation of the algorithm to compute the nullspace, using the 
  \textit{SciPy} library \cite{scipy/ranknullspace}:

  \begin{lstlisting}[language=python]
def nullspace(A, atol=1e-13, rtol=0):
    A = np.atleast_2d(A)
    u, s, vh = svd(A)
    tol = max(atol, rtol * s[0])
    nnz = (s >= tol).sum()
    ns = vh[nnz:].conj().T
    return ns
    \end{lstlisting}
\end{exmp}

@@
